A Standardized Effect Size for Evaluating the Comparing the Strength of Phylogenetic Signal

### 2020-07-15 Manuscript/OldVsns/Manuscript-4-EKB.Rmd
pnas vsn

It seems intuitive to interpret the strength of phylogenetic signal based on the value of $\lambda$, as $\lambda$ is a parameter on a bounded scale ($0\to1$) for which interpretation of its extremal points are understood. Specifically, $\lambda=0$ represents no phylogenetic signal, while $\lambda=1$ is phylogenetic signal as expected under Brownian motion. However, equating values of $\lambda$ directly to the strength of phylogenetic signal presumes two important statistical properties that have not been fully explored. First, it presumes that values of $\lambda$ can be precisely estimated, as biological inferences regarding the strength of phylogenetic signal depend on high accuracy in its estimation. Therefore, understanding the precision in estimating $\lambda$ is paramount. One study [@Boettiger_et_al2012] found that estimates of Pagel's $\lambda$ displayed less variation (i.e., greater precision) when data were simulated on a large phylogeny ($N=281$) as compared to a small one ($N=13$). From this observation it was concluded that insufficient data (i.e., the number of species) was the underlying cause of the increased variation across parameter estimates [@Boettiger_et_al2012]. Indeed, such a pattern is common with statistical estimators, as summary statistics and parameters are often more precise at greater sample sizes [@Cohen1988]. However, this conclusion also implies that the precision of $\lambda$ remains constant across its range ($\lambda = 0 \to 1$); an assumption that to date, has not been verified. Thus, despite widespread use of Pagel's [-@Pagel1999] $\lambda$ in macroevolutionary studies, at present, we lack a general understanding of the precision with which $\lambda$ can estimate levels of phylogenetic signal in phenotypic datasets. \hfill\break

Second, while estimates of $\lambda$ are within a bounded scale ($0\to1$), this does not *de-facto* imply that the estimated values of this parameter correspond to the actual strength of the underlying input signal in the data. For this to be the case, $\lambda$ must be a statistical effect size. Effect sizes are a measure the magnitude of a statistical effect in data, represented on a common scale [@Glass1976; @Cohen1988]. Effect sizes have widespread use in many areas of the quantiative sciences, as they represent measures that may be readily summarized across datasets as in meta-analyses [@HedgesOlkin1985; @Glass1976; @Arnqvist1995], or compared among datasets [e.g., @AdamsCollyer2016; @AdamsCollyer2019b]. Unfortunatley, not all model parameters and test statistics are effect sizes, and thus many summary measures must first be converted to standardized units (i.e., an effect size) for meaningful comparison [see @Rosenthal1994]. As a consequence, it follows that only if $\lambda$ is a statistical effect size can comparisons of estimates across datasets be interpretable. For the case of $\lambda$, this has not yet been explored.  \hfill\break

@@

*Results.* We found that the precision of $\lambda_{est}$ varied widely across simulation conditions. Predictably, precision improved as the number of species increased (Figure 2). This confirmed earlier findings of Boettiger et al. [-@Boettiger_et_al2012], and adhered to parametric statistical theory. However, in many cases the set of $\lambda_{est}$ spanned nearly the entire range of possible values (e.g., $n=32$; $\lambda_{in}=0.5$: $\lambda_{est}= 0.0\to 0.985$), revealing that estimates of $\lambda$ were not a reliable indicator of input phylogenetic signal. Importantly, the precision of $\lambda_{est}$ was not uniform across all levels of phylogenetic signal. The worst precision was observed at intermediate levels of phylogenetic signal ($\lambda_{in}\approx0.5$), while precision improved as input levels approached the extremes of $\lambda$'s range (i.e., $\lambda_{in}\rightarrow0$ & $\lambda_{in}\rightarrow1$). Thus, estimates of $\lambda$ were least reflective of the true input signal at intermediate values. Additionally, even at large levels of species richness, we found that the range of $\lambda_{est}$ still encompassed a substantial portion of possible values (e.g., $n=512$; $\lambda_{in}=0.5$: $\lambda_{est} = 0.32\to 0.68$). Likewise, the same $\lambda_{est}$ could be obtained from datasets containing vastly different input levels of phylogenetic signal (e.g., $n=512$; $\lambda_{est} = 0.5$; $\lambda_{in} = 0.25 \to 0.65$). These findings were particularly unsettling when considered in light of our literature survey. Over one quarter of the $\lambda$ estimates obtained in empirical studies (421 of 1,572) were between $\lambda=0.25$ and $\lambda=0.75$ (Figure 1). This range reflected the region that our simulations identified as being the least reliable in terms of accurately characterizing levels of phylogenetic signal, yet 30% of these mid-range empirical estimates were explicitly interpreted in terms of the strength of phylogenetic signal that they represented (i.e., weak, intermediate, strong phylogenetic signal). \hfill\break

@@

To evaluate the utility of $Z_K$ and $Z_\lambda$ we calculated both effect sizes for the simulated datasets generated above, and summarized the precision of each using its variance ($\sigma^2_{Z_K}$ and $\sigma^2_{Z_\lambda}$). Results are found in Figure 4 (additional results are found in the Supporting Information). Here two things are evident. First, estimates of $Z_K$ track the input phylogenetic signal in a more linear fashion than do estimates of $Z_\lambda$ (Figure 4A,B). Thus, actual changes in the strength of phylogenetic signal are reflected more evenly in the corresponding values of the effect size $Z_K$. Second, the precision of $Z_K$ is considerably more stable as compared with $Z_\lambda$. This may be seen by calculating the coefficients of variation for the set of precision estimates (i.e., $\sigma^2_{Z_K}$ and $\sigma^2_{Z_\lambda}$) across input levels of phylogenetic signal.  Here coefficients of variation in the precision of $Z_K$ were an order of magnitude smaller for than for $Z_\lambda$ (Figure 4C). This implied that estimates of the strength of phylogenetic signal were more reliable and robust when using $Z_K$ as compared with $Z_\lambda$. \hfill\break

@@

Once the magnitude of phylogenetic signal is characterized using $Z_K$, one may wish to compare such measures across datasets, to determine whether the strength of phylogenetic signal is significantly greater in one phenotypic trait as compared to another. As with other effect sizes derived from permutation distributions [e.g., @AdamsCollyer2016; @AdamsCollyer2019b], a two-sample test statistic may be calculated as:

@@

To demonstrate the utility of $\hat{Z}_{12}$ we performed an analysis of the strength of phylogenetic signal in two phenotypic datasets from species of plethodontid salamander. The data were part of a series of studies examining macroevolutionary trends in phenotypic diversification in this group [@Baken2019; @Baken2020]. Our dataset contained surface area to volume ratios (SA:V) and relative body width ($\frac{BW}{SVL}$) for 305 species (Figure 5A). For SA:V, 11 linear body measurements from 2,781 individuals were taken, from which estimates of the surface area and volume of the head, body, and tail were calculated and subsequently combined to arrive at the SA:V for each indiviual [for mathematical details see @Baken2020]. Species means were then obtained. Likewise, body size (SVL) and body width (BW) measurements were taken from 3,371 individuals, and species means of relative body width ($\frac{BW}{SVL}$) were calculated [data from @Baken2019]. A time-dated molecular phylogeny for the group [@Bonett2017] was then pruned to match the species in the dataset, resulting in a phylogeny and corresponding phenotypic dataset containing 305 species. The phylogenetic signal in each trait was then characterized using $Kappa$, which was converted to its effect size ($Z_K$) using `geomorph` 3.2.1 [@AdamsOtarola2013; @AdamsGeomorph]. Finally, the strength of phylogenetic signal was compared across traits using $\hat{Z}_{12}$ as described above (**to be incorporated in `geomorph` upon manuscript acceptance**). \hfill\break

@@

As an alternative, we described a standardized effect size ($Z$) for assessing the strength of phylogenetic signal. $Z$ expresses the magnitude of phylogenetic signal as a standard normal deviate, which is easily interpretable as the strength of phylogenetic signal relative to the mean. We applied this concept to both $\lambda$ and $Kappa$, and found that $Z_K$ was a better estimate of the strength of phylogenetic signal in phenotypic data. First, $Z_K$ was more precise than $Z_{\lambda}$, and variation in this precision was more consistent across the range of input levels of phylogenetic signal. Additionally, values of $Z_K$ more accurately tracked known levels of phylogenetic signal, with changes in the actual strength of phylogenetic signal reflected in a more linear fashion by concomite changes in the values of $Z_K$. Thus, $Z_K$ holds promise as a measure of the relative strength of phylogenetic signal that reflects the magnitude of this effect in phenotypic data. We therefore recommend that future studies interested in the strength of phylogenetic signal incorporate $Z_K$ as a statistical measure of this effect. \hfill\break

@@

One important advantage of the approach advocated here is that the resulting effect sizes ($Z_K$) are dimensionless, as the units of measurement cancel out during the calculation of $Z$ [@SokalRohlf2012]. Thus, $Z_K$ represents the strength of phylogenetic signal on a common and comparable scale -- measured in standard deviation units -- regardless of the initial units and original scale of the phenotypic variables under investigation. This means that the strength of phylogenetic signal may be compared across datsets for continuous phenotypic traits measured in different units and scale, because those units have been standardized through their conversion to $Z_K$. For example, our approach could be utilized to determine whether the strength of phylogenetic signal (say, in response to ecological differentiation) is stronger in morphological traits (linear traits: $mm$), physiological traits (metabolic rate: $\frac{O^2}{min}$), or behavioral traits (aggression: $\frac{\#{displays}}{second}$). Additionally, our method  is capable of comparing the strength of phylogenetic signal in traits of different dimensionality, as estimates of phylogenetic signal using $Kappa$ have been generalized for multivariate data [$K_{mult}$: see @Adams2014a]. Likewise, as $Z_K$ is a standard deviation unit (which accounts for variation, sample size, and the phylogeny), tests based on $\hat{Z}_{12}$ may be utilized for comparing the strength of phylogenetic signal across datasets containing a different number of species, and even for phenotypes obtained from species in different lineages, because their phylogenetic non-independence, and observed variation, are taken into account in the generation of the empirical sampling distribution via permutation.    \hfill\break

>>>>

It seems intuitive to interpret the strength of phylogenetic signal based on the value of $\lambda$, as $\lambda$ is a parameter on a bounded scale ($0\to1$) for which interpretation of its extremal points are understood. Specifically, $\lambda=0$ represents no phylogenetic signal, while $\lambda=1$ is phylogenetic signal as expected under Brownian motion, However, equating values of $\lambda$ directly to the strength of phylogenetic signal presumes ~~two important statistical properties that have not been fully explored. First, it presumes that values of $\lambda$ can be precisely estimated, as biological inferences regarding the strength of phylogenetic signal depend on high accuracy in its estimation. Therefore, understanding the precision in estimating $\lambda$ is paramount.~~ THE IMPORTANT STATISTICAL PROPERTY OF PRECISION. One study BRIEFLY EXPLORED THIS CONCEPT [@Boettiger_et_al2012] AND found that estimates of Pagel's $\lambda$ displayed less variation (i.e., greater precision) when data were simulated on a large phylogeny ($N=281$) as compared to a small one ($N=13$). From this observation, it was concluded that insufficient data (i.e., the number of species) was the underlying cause of the increased variation across parameter estimates [@Boettiger_et_al2012]. Indeed, such a pattern is common with statistical estimators, as summary statistics and parameters are often more precise at greater sample sizes [@Cohen1988]. However, this conclusion also implies that the precision of $\lambda$ remains constant across its range ($\lambda = 0 \to 1$); an assumption that to date, has not been verified. Thus, despite widespread use of Pagel's [-@Pagel1999] $\lambda$ in macroevolutionary studies, at present, we lack a general understanding of the precision with which $\lambda$ can estimate levels of phylogenetic signal in phenotypic datasets.~~ [EKB: I suggested the rewrite above because I feel like the two presumptions of precision and consistent precision across the range can be simplified. As always, feel free to ignore what you dislike.]   \hfill\break

Second, while estimates of $\lambda$ are within a bounded scale ($0\to1$), this does not *de-facto* imply that the estimated values of this parameter correspond to the actual strength of the underlying input signal in the data. For this to be the case, $\lambda$ must be a statistical effect size. Effect sizes are a measure OF the magnitude of a statistical effect in data, represented on a common scale [@Glass1976; @Cohen1988]. Effect sizes have widespread use in many areas of the quantiative sciences, as they represent measures that may be readily summarized across datasets as in meta-analyses [@HedgesOlkin1985; @Glass1976; @Arnqvist1995], or compared among datasets [e.g., @AdamsCollyer2016; @AdamsCollyer2019b]. Unfortunatley, not all model parameters and test statistics are effect sizes, and thus many summary measures must first be converted to standardized units (i.e., an effect size) for meaningful comparison [see @Rosenthal1994]. As a consequence, it follows that only if $\lambda$ is CONVERTED TO a statistical effect size can comparisons of estimates across datasets be interpretable. For the case of $\lambda$, this has not yet been explored.  \hfill\break

@@

*Results.* We found that the precision of $\lambda_{est}$ varied widely across simulation conditions. Predictably, precision improved as the number of species increased (Figure 2). This confirmed earlier findings of Boettiger et al. [-@Boettiger_et_al2012], and adhered to parametric statistical theory. However, in many cases the set of $\lambda_{est}$ spanned nearly the entire range of possible values (e.g., $n=32$; $\lambda_{in}=0.5$: $\lambda_{est}= 0.0\to 0.985$), revealing that estimates of $\lambda$ were not a reliable indicator of input phylogenetic signal. Importantly, the precision of $\lambda_{est}$ was not uniform across all levels of phylogenetic signal. The worst precision was observed at intermediate levels of phylogenetic signal ($\lambda_{in}\approx0.5$), while precision improved as input levels approached the extremes of $\lambda$'s range (i.e., $\lambda_{in}\rightarrow0$ & $\lambda_{in}\rightarrow1$). Thus, estimates of $\lambda$ were least reflective of the true input signal at intermediate values. Additionally, even at large levels of species richness, we found that the range of $\lambda_{est}$ still encompassed a substantial portion of possible values (e.g., $n=512$; $\lambda_{in}=0.5$: $\lambda_{est} = 0.32\to 0.68$). Likewise, the same $\lambda_{est}$ could be obtained from datasets containing vastly different input levels of phylogenetic signal (e.g., $n=512$; $\lambda_{est} = 0.5$; $\lambda_{in} = 0.25 \to 0.65$). These findings were particularly unsettling when considered in light of our literature survey [WE DON'T EXPLICITLY SAY THAT WE CONDUCTED A LITERATURE SURVEY BEFORE THIS POINT, JUST MENTION SOME RESULTS FROM IT IN THE INTRO... IS THAT OK OR COULD WE ADD SOME PHRASING TO THAT SECTION OF THE INTRO SO AT LEAST THE PHRASE 'LITERATURE SURVEY' APPEARS BEFORE HERE?]. Over one quarter of the $\lambda$ estimates ~~obtained~~ PUBLISHED in empirical studies (421 of 1,572) were between $\lambda=0.25$ and $\lambda=0.75$ (Figure 1). This range reflected the region that our simulations identified as being the least reliable in terms of accurately characterizing levels of phylogenetic signal, yet 30% of these mid-range empirical estimates were explicitly interpreted in terms of the strength of phylogenetic signal that they represented (i.e., weak, intermediate, strong phylogenetic signal). \hfill\break

@@

To evaluate the utility of $Z_K$ and $Z_\lambda$ we calculated both effect sizes for the simulated datasets generated above, and summarized the precision of each using its variance ($\sigma^2_{Z_K}$ and $\sigma^2_{Z_\lambda}$~~). Results are found in Figure 4 (~~, FIGURE 4, additional results ~~are found~~ in the Supporting Information). Here, two things are evident. First, estimates of $Z_K$ LINEARLY track the input phylogenetic signal ~~in a more linear fashion than do~~ WHEREAS estimates of $Z_\lambda$ DO NOT (Figure 4A,B). Thus, actual changes in the strength of phylogenetic signal are reflected more evenly in the corresponding values of the effect size $Z_K$. Second, the precision of $Z_K$ is considerably more stable as compared with $Z_\lambda$. This may be seen by calculating the coefficients of variation for the set of precision estimates (i.e., $\sigma^2_{Z_K}$ and $\sigma^2_{Z_\lambda}$) across input levels of phylogenetic signal. Here coefficients of variation in the precision of $Z_K$ were an order of magnitude smaller for than for $Z_\lambda$ (Figure 4C). This implied that estimates of the strength of phylogenetic signal were more reliable and robust when using $Z_K$ as compared with $Z_\lambda$. \hfill\break

@@

Once the magnitude of phylogenetic signal is characterized using $Z_K$, one may wish to compare such measures across datasets, to determine whether the strength of phylogenetic signal is significantly greater in one phenotypic trait ~~as compared to~~ THAN another [EKB: changed wording to avoid repetition of word "compare"]. As with other effect sizes derived from permutation distributions [e.g., @AdamsCollyer2016; @AdamsCollyer2019b], a two-sample test statistic may be calculated as:

@@

To demonstrate the utility of $\hat{Z}_{12}$ we ~~performed an analysis of~~ QUANTIFIED AND COMPARED the strength of phylogenetic signal ~~in~~ OF two phenotypic ~~datasets~~ TRAITS WITH DIFFERENT UNITS ACROSS ~~from~~ species of plethodontid salamander. The data were part of a series of studies examining macroevolutionary trends in phenotypic diversification in this group [@Baken2019; @Baken2020]. Our dataset contained surface area to volume ratios (SA:V; $mm^{-1}$) and relative body width ($\frac{BW}{SVL}$; NO UNITS) for 305 species (Figure 5A). For SA:V, 11 linear body measurements from 2,781 individuals were taken, from which estimates of the surface area and volume of the head, body, and tail were calculated and subsequently combined to arrive at the SA:V for each indiviual [for mathematical details see @Baken2020]. Species means were then obtained. Likewise, body size (SVL) and body width (BW) measurements were taken from 3,371 individuals, and species means of relative body width ($\frac{BW}{SVL}$) were calculated [data from @Baken2019]. A time-dated molecular phylogeny for the group [@Bonett2017] was then pruned to match the species in the dataset, resulting in a phylogeny and corresponding phenotypic dataset containing 305 species. The phylogenetic signal in each trait was then characterized using $Kappa$, which was converted to its effect size ($Z_K$) using `geomorph` 3.2.1 [@AdamsOtarola2013; @AdamsGeomorph]. Finally, the strength of phylogenetic signal was compared across traits using $\hat{Z}_{12}$ as described above (**to be incorporated in `geomorph` upon manuscript acceptance**). \hfill\break

@@

As an alternative, we described a standardized effect size ($Z$) for assessing the strength of phylogenetic signal. $Z$ expresses the magnitude of phylogenetic signal as a standard normal deviate, which is easily interpretable as the strength of phylogenetic signal relative to the mean. We applied this concept to both $\lambda$ and $Kappa$, and found that $Z_K$ was a better estimate of the strength of phylogenetic signal in phenotypic data. First, $Z_K$ was more precise than $Z_{\lambda}$, and variation in this precision was more consistent across the range of input levels of phylogenetic signal. Additionally, values of $Z_K$ more accurately tracked known levels of phylogenetic signal, with changes in the actual strength of phylogenetic signal reflected in a more linear fashion by ~~concomite~~ CONCOMITANT changes in the values of $Z_K$. Thus, $Z_K$ holds promise as a measure of the relative strength of phylogenetic signal that reflects the magnitude of this effect in phenotypic data. We therefore recommend that future studies interested in the strength of phylogenetic signal incorporate $Z_K$ as a statistical measure of this effect. \hfill\break

@@

One important advantage of the approach advocated here is that the resulting effect sizes ($Z_K$) are dimensionless, as the units of measurement cancel out during the calculation of $Z$ [@SokalRohlf2012]. Thus, $Z_K$ represents the strength of phylogenetic signal on a common and comparable scale -- measured in standard deviation units -- regardless of the initial units and original scale of the phenotypic variables under investigation. This means that the strength of phylogenetic signal may be compared across datsets for continuous phenotypic traits measured in different units and scale because those units have been standardized through their conversion to $Z_K$. For example, our approach could be utilized to determine whether the strength of phylogenetic signal (say, in response to ecological differentiation) is stronger in morphological traits (linear traits: $mm$), physiological traits (metabolic rate: $\frac{O^2}{min}$), or behavioral traits (aggression: $\frac{\#{displays}}{second}$). Additionally, our method  is capable of comparing the strength of phylogenetic signal in traits of different dimensionality, as estimates of phylogenetic signal using $Kappa$ have been generalized for multivariate data [$K_{mult}$: see @Adams2014a]. Likewise, as $Z_K$ is a standard deviation unit (which accounts for variation, sample size, and the phylogeny), tests based on $\hat{Z}_{12}$ may be utilized for comparing the strength of phylogenetic signal across datasets containing a different number of species, ~~and~~ even for phenotypes obtained from species in different lineages because their phylogenetic non-independence, and observed variation, are taken into account in the generation of the empirical sampling distribution via permutation.    \hfill\break
